Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2019-03-26 RNARedPrint.tex
Fix remaining todos for R2

\usepackage{todonotes}

@@

Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions
associated with local structural motifs (base pairs, base pair stacks, loops, \ldots) under a certain nucleotide
assignment. Energy models can thus be captured generically by introducing, for each motif $\motif$ occurring in
a target structure, a contribution $f_{\motif}$, taking a specific value for each assignment of nucleotides to its
positions $\dep(f_{\motif})$. For instance, the contribution of a \Def{base pair stack}, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, can be captured by the introduction of a function $f^{\textsf{Stack}}_{\substack{i,j}}$ such that $\dep(f^{\textsf{Stack}}_{\substack{i,j}}) = \{i,i+1,j-1,j\}$.
We refer to energy models that consider the contributions of all base pair stacks (and thus introduce the corresponding dependencies) collectively as the \Def{stacking energy model}.
}

@@

%For this purpose, we define the \Def{transformed contribution} 
%$\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
%such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$ 
%
Let us now specify the concrete set $\F$ of contributions that we use
for the design in the base pair model\todo{Didn't we plan to get rid of the bp model?}{} targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$. The set $\F$ thus consists of
\begin{itemize}
\item the transformed contributions $\ln(\pi_0)f^{\GCb}_i$ for the ~$\GCb\%$ feature ($i=1,\dots,n$);
\item the transformed contributions $\ln(\pi_\ell)f^{R_\ell}_{ij}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
\end{itemize}

By these definitions, the set $\F$ encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}).
%Subsequently, we show how to efficiently compute this partition function based on the 
%corresponding dependency graph and its tree decomposition.

@@

\begin{theorem}[Complexities]\label{th:complexities}
  For sequence length $n$, $k$ target structures and treewidth $w$,
  % and a base pair dependency graph having $c$ connected components%
  $t$ sequences
  are generated from the Boltzmann distribution  in
  $O( n\, k \, 4^{w+1} + t\, n\, k )$ time.
  %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
  %$\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
\end{theorem}

@@

In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on more expressive energy models \revisedOne{or 2) by sampling in a simple energy model which} can be used to approximate sampling in more complex models.  In practice, complex energy models \revisedOne{have} a strong influence on the treewidth (of optimal tree decompositions) of the dependency graph and \revisedOne{thus on} the computational complexity of our approach. Therefore, it is interesting to consider---in addition to the base pair model---other
stripped-down variants of the nearest neighbor model, which could offer a compromise between the low-complexity base pair energy model and the high-accuracy Turner model.

@@

\begin{proposition}\label{prop:bijection}
  $|\Design{G}| = |\IS{G}|$.
\end{proposition}
%
\begin{proof}
Consider the mapping $\Psi: \Design{G} \to \IS{G}, f \mapsto \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$
We show that $\Psi$ is bijective:%
\begin{itemize}
\item $\Psi$ is injective, i.e. $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
If $f\neq f'$, then there exists a node $v\in V$ such that $f(v)\neq f'(v)$.
We discuss only the case $v\in V_1$, where we restricted the nucleotides to $\Ab$ and $\Gb$. Then, $\{f(v),f'(v)\}$ must equal $\{\Ab,\Gb\}$, such that either $v\in\Psi(f)$ or $v\in\Psi(f')$.
%
\item $\Psi$ is surjective, i.e. there is a preimage for each element $I\in \IS{G}$. Define $f\in\Design{G}$ as\begin{displaymath}
f(v) = \left\{\begin{array}{ll} % @{\qquad}ll
\Ab{} & \text{if $v\in V_1$ and $v\in I$} \\ %&
\Cb{} & \text{if $v\in V_2$ and $v\in I$} \\
\Gb{} & \text{if $v\in V_1$ and $v\not\in I$} \\ %&
\Ub{} & \text{if $v\in V_2$ and $v\not\in I$}
\end{array}\right.
\end{displaymath}
One easily verifies that $\Psi(f) = I$.
%
%for any $(v_1,v_2)\in V_1\times V_2$ by
%\begin{align*}
%f(v_1) &= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} &
%\text{\quad and\quad}&&
% f(v_2) &= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S.\end{cases}
%\end{align*}
%
It remains to show that $f$ is a valid design for $G$, i.e. for each $(v, v') \in E$, $\{f(v),f(v')\}\in \B$; please recall that we defined $\B$ as the set of all valid nucleotide pairs.
Assume there is an edge $(v_1,v_2)\in E$, violating $\{f(v_1),f(v_2)\} \in \B$. Since $G$ is bipartite,
$v_1\in V_1$ and $v_2\in V_2$, such that $f(v_1)\in\{\Ab,\Gb\}$ and $f(v_2)\in\{\Cb,\Ub\}$. This implies that among all possible $\{f(v_1), f(v_2)\}$ only $\{\Ab,\Cb\}$ is not in $\B$, which in turn requires $v_1\in I$ and $v_2\in I$. Therefore, since $I$ is an independent set, the edge $(v_1,v_2)\in E$ cannot exist. \QEDA
\end{itemize}
\end{proof}

@@

We implemented the core algorithms in \texttt{C++}, resulting in the tool
\ourprog{}, available at:

{\centering \url{https://github.com/yannponty/RNARedPrint}\\[.3em]}%

\ourprog{}  takes a set of target structures, as well as
weights for each energy feature and \GCb\%, and generates a sample set of sequences compatible with the structures in the corresponding
Boltzmann distribution. \ourprog{} currently supports the base pair
energy model and the stacking energy model.

Moreover, we provide two {\tt Python} wrapper scripts. 
The first script targets prescribed energies using multi-dimensional Boltzmann sampling.
For a given set of secondary structures, together with prescribed target energies
and target \GCb\%, this script generates a series of sequences that satisfy the target values for the energy and
\GCb\% features within configurable tolerances.
Notably, these target energies are actual free energies in the realistic
Turner energy model, and are targeted by efficiently sampling in the
base pair energy model, and filtering sequences based on the {\tt RNAeval} tool from the Vienna package~\cite{Lorenz2011}. The second script generates high quality seed sequences suitable for negative RNA design. The details of this approach are described in the subsections below.

@@

\section*{Additional Files}

\subsection*{Additional file 1}
\revised{Supplemental Material providing additional detailed information on 
``Immediate benefits of positive design for negative design'', ``Tree decomposition for RNA design instances in
practice'', ``Run times for Boltzmann sampling (at fixed weights)'', ``Illustrating multi-dimensional Boltzmann sampling
for three pseudoknot-free target structures'', ``Effectivity of the seed sequence generation'', ``Parameters for the
base pair and the stacking energy model'', ``Tree decomposition with hyper-dependencies'', ``Correctness of the FPT
partition function algorithm'', ``General complexity of the partition function computation by cluster tree elimination
and generation of samples'', ``Exploiting constraint consistency to reduce the complexity'', ``Monotonicity of the
partial derivatives within weight calibration'', ``Approximate counting and random generation'', ``Contribution of terms
to the MultiDefect score'', and ``Detailed results of the negative design benchmark''.}

@@

\section*{Availability of data and materials}
The software and data of this work are available in the Github repository \url{https://github.com/yannponty/RNARedPrint}.
\todo[inline]{This section should include a link to the most recent version of your software or code (e.g. GitHub or
  Sourceforge) as well as a link to the archived version referenced in the manuscript. The software or code should be
  archived in an appropriate repository with a DOI or other unique identifier. For software in GitHub, we recommend
using Zenodo.}

>>>>

@@

Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions
associated with local structural motifs (base pairs, base pair stacks, loops, \ldots) under a certain nucleotide
assignment. Energy models can thus be captured generically by introducing, for each motif $\motif$ occurring in
a target structure, a contribution $f_{\motif}$, taking a specific value for each assignment of nucleotides to its
positions $\dep(f_{\motif})$. For instance, the contribution of a \Def{base pair stack}, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, can be captured by the introduction of a function $f^{\textsf{Stack}}_{\substack{i,j}}$ such that $\dep(f^{\textsf{Stack}}_{\substack{i,j}}) = \{i,i+1,j-1,j\}$.
We refer to energy models that consider the contributions of all base pair stacks (and thus introduce the corresponding
dependencies) collectively as the \Def{stacking energy model} \revised{(briefly, \Def{stacking model})}.
}

@@

%For this purpose, we define the \Def{transformed contribution} 
%$\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
%such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$ 
%
Let us now specify the concrete set $\F$ of contributions that we use
for the design in the stacking energy model targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$. The set $\F$ thus consists of
\begin{itemize}
\item the transformed contributions $\ln(\pi_0)f^{\GCb}_i$ for the ~$\GCb\%$ feature ($i=1,\dots,n$);
\item the transformed contributions $\ln(\pi_\ell)f^{{\textsf{Stack}}}_{ij}$ for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
\end{itemize}
%
By these definitions, the set $\F$ encodes the partition function $\partfun{\pi_0,\dots,\pi_k}$ of Eq.~(\ref{eq:mainproblem}).
%Subsequently, we show how to efficiently compute this partition function based on the 
%corresponding dependency graph and its tree decomposition.

@@

\begin{theorem}[Complexities]\label{th:complexities}
  \revised{Given are sequence length $n$, $k$ target structures, and tree\-width $w$.
  % and a base pair dependency graph having $c$ connected components%
  $t$ sequences
  are generated from the Boltzmann distribution  in
  $O( n\, k \, 4^{w+1} + t\, n\, k )$ time.}
  %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
  %$\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
\end{theorem}

@@

In order to capture realistic energy models like the Turner model or pseudoknot models like {\tt
HotKnots}~\citep{Ren2005}, our sampling strategy can be extended in two ways: 1) either by directly sampling based on
more expressive energy models \revisedOne{or 2) by sampling in a simple energy model which} can be used to approximate
sampling in more complex models.  In practice, complex energy models \revisedOne{have} a strong influence on the
treewidth (of optimal tree decompositions) of the dependency graph and \revisedOne{thus on} the computational complexity
of our approach. Therefore, it is interesting to consider---in addition to the \revised{stacking energy model}---other
stripped-down variants of the nearest neighbor model, which could offer a compromise between \revised{low-complexity (as due to
the stacking energy model) and the high-accuracy of the Turner model.}

@@

\begin{proposition}\label{prop:bijection}
  $|\Design{G}| = |\IS{G}|$.
\end{proposition}
%
\begin{proof}
  Consider the mapping\\\hspace*{2em}$\Psi: \Design{G} \to \IS{G}, f \mapsto \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$\\
We show that $\Psi$ is bijective:%
\begin{itemize}
\item $\Psi$ is injective, i.e. $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
If $f\neq f'$, then there exists a node $v\in V$ such that $f(v)\neq f'(v)$.
We discuss only the case $v\in V_1$, where we restricted the nucleotides to $\Ab$ and $\Gb$. Then, $\{f(v),f'(v)\}$ must equal $\{\Ab,\Gb\}$, such that either $v\in\Psi(f)$ or $v\in\Psi(f')$.
%
\item $\Psi$ is surjective, i.e. there is a preimage for each element $I\in \IS{G}$. Define $f\in\Design{G}$ as\begin{displaymath}
f(v) = \left\{\begin{array}{ll} % @{\qquad}ll
\Ab{} & \text{if $v\in V_1$ and $v\in I$} \\ %&
\Cb{} & \text{if $v\in V_2$ and $v\in I$} \\
\Gb{} & \text{if $v\in V_1$ and $v\not\in I$} \\ %&
\Ub{} & \text{if $v\in V_2$ and $v\not\in I$}
\end{array}\right.
\end{displaymath}
One easily verifies that $\Psi(f) = I$.
%
%for any $(v_1,v_2)\in V_1\times V_2$ by
%\begin{align*}
%f(v_1) &= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases} &
%\text{\quad and\quad}&&
% f(v_2) &= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S.\end{cases}
%\end{align*}
%
It remains to show that $f$ is a valid design for $G$, i.e. for each $(v, v') \in E$, $\{f(v),f(v')\}\in \B$; please recall that we defined $\B$ as the set of all valid nucleotide pairs.
Assume there is an edge $(v_1,v_2)\in E$, violating $\{f(v_1),f(v_2)\} \in \B$. Since $G$ is bipartite,
$v_1\in V_1$ and $v_2\in V_2$, such that $f(v_1)\in\{\Ab,\Gb\}$ and $f(v_2)\in\{\Cb,\Ub\}$. This implies that among all possible $\{f(v_1), f(v_2)\}$ only $\{\Ab,\Cb\}$ is not in $\B$, which in turn requires $v_1\in I$ and $v_2\in I$. Therefore, since $I$ is an independent set, the edge $(v_1,v_2)\in E$ cannot exist. \QEDA
\end{itemize}
\end{proof}

@@

We implemented the core algorithms in \texttt{C++}, resulting in the tool
\ourprog{}, available at: 
%
{\url{https://github.com/yannponty/RNARedPrint}}.%

\noindent
\ourprog{}  takes a set of target structures, as well as
weights for each energy feature and \GCb\%, and generates a sample set of sequences compatible with the structures in the corresponding
Boltzmann distribution\revised{; it currently supports the stacking energy model and the base pair energy model
(Additional file~1: Section~\ref{appsec:modelparameters}).}

Moreover, we provide two {\tt Python} wrapper scripts. 
The first script targets prescribed energies using multi-dimensional Boltzmann sampling.
For a given set of secondary structures, together with prescribed target energies
and target \GCb\%, this script generates a series of sequences that satisfy the target values for the energy and
\GCb\% features within configurable tolerances.
Notably, these target energies are actual free energies in the realistic
Turner energy model, and are targeted by efficiently sampling in the
stacking energy model, and filtering sequences based on the {\tt RNAeval} tool from the Vienna package~\cite{Lorenz2011}. The second script generates high quality seed sequences suitable for negative RNA design. The details of this approach are described in the subsections below.

@@

\section*{Additional file}

\subsection*{Additional file 1}
\revised{The Supplemental Material contains additional
information on methods and parameters; elaboration of theory and proofs; and additional and detailed empirical results.}
% 
% \revised{Supplemental Material providing additional detailed information on 
% ``Immediate benefits of positive design for negative design'', ``Tree decomposition for RNA design instances in
% practice'', ``Run times for Boltzmann sampling (at fixed weights)'', ``Illustrating multi-dimensional Boltzmann sampling
% for three pseudoknot-free target structures'', ``Effectivity of the seed sequence generation'', ``Parameters for the
% base pair and the stacking energy model'', ``Tree decomposition with hyper-dependencies'', ``Correctness of the FPT
% partition function algorithm'', ``General complexity of the partition function computation by cluster tree elimination
% and generation of samples'', ``Exploiting constraint consistency to reduce the complexity'', ``Monotonicity of the
% partial derivatives within weight calibration'', ``Approximate counting and random generation'', ``Contribution of terms
% to the MultiDefect score'', and ``Detailed results of the negative design benchmark''.}

@@

\section*{Availability of data and materials}
The software and data of this work are maintained in the Github repository
\url{https://github.com/yannponty/RNARedPrint}\revised{; the acrchived version referenced in this manuscript is
available at \href{https://dx.doi.org/10.5281/zenodo.2597571}{doi:10.5281/zenodo.2597571}.}
